---
title: "[Assess] Bracero Farmworker Case Study"
subtitle: "📍 Yakima Valley, WA"
date: "September 19, 2025"
format: 
  revealjs:
    theme:
      - default
    css: custom.scss
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    center: true
execute:
  freeze: auto
---

# Class Overview 
- Centering activity
- Setting the stage
- Paired coding
- Group Debrief

# Centering Activity

# Setting the Stage
- This week's focus is on data tidying and transformation
- Using metadata from 90 interviews in the Bracero Archive that I scraped
- Important to note that each row represents a person, a life, and a family impacted by the Bracero Program

# Let's get Started
- Find your pair
- Open Posit

# First Steps
- Import Data
- Load packages
```{r}
#| warning: false
#| message: false
#| echo: true

library(readxl) # for loading data
library(tidyverse) # for tidying
```

```{r}
# | eval: true
# | echo: false
bracero <- read_excel("~/Google Drive/My Drive/Teaching/ES-202/Case Study Code/data/week_03/bracero_pull.xlsx")
```

# Load Data
```{r}
#| warning: false
#| message: false
#| echo: true
#| eval: false
bracero <- read_excel("data/bracero_pull.xlsx")
```

# Part II: Today's Goal
- Understand data structure
- Transform and tidy data for necessary questions
- Explore and summarize patterns
- Connect it to theoretical understandings of procedural justice

## Our data {.smaller}
- 90 rows (interviews)
- 6 variables (all `chr` type)
  - title = Person interviewed
  - subject = topic of interview
  - description = summary of interview
  - date = date of interview
  - identifier = link to original interview file
  - language = language interview was conducted in
```{r}
#| warning: false
#| message: false
#| echo: true
glimpse(bracero)
```

## Data Types {.smaller}
- `chr` is character
- `dbl` is double (number)
- `int` is an integer
- `dttm` is date-time 
- `fct` is factor

<span style="color:red;">Thoughts on what variables we may want to change the data types of?</span>

```{r}
#| warning: false
#| message: false
#| echo: false

glimpse(bracero)
```

# Transforming Data (Columns) {.smaller}
- Four functions can change the columns of your data without affecting your rows 
  - mutate(): creates new columns based on existing ones
  - select(): chooses what columns you want in your dataframe
  - rename(): changes the names of current columns
  - relocate(): changes the position of those columns 

## mutate() function {.smaller}
- We use the mutate function to change the `date` column to a date format. 

```{r}
#| warning: false
#| message: false
#| echo: true

bracero |>
  mutate(date_format = as.Date(date)) 
```

## mutate() function (cont.) {.smaller}
- We can also use mutate for creating new columns, and we can pair it with additional functions, such as ifelse statements. 
- Perhaps we were interested in creating a new column that defines whether an interview is from a bracero worker or not. 
  - Note that if we wanted to further classify the subject, we might create a factor column and use the `case_match()` function, rather than an ifelse statement.
```{r}
#| warning: false
#| message: false
#| echo: true 

bracero |> 
  mutate(subject_binary = ifelse(subject %in% c("bracero", "Bracero"), "Bracero", "Non-Bracero" )) 
```

## select() function {.smaller}
- Perhaps we'd like to remove the original `date` column from our dataframe. 
- `select()` can be used to select or deselect columns. 
```{r}
#| warning: false
#| message: false
#| echo: true
#| eval: false
#| code-line-numbers: "|3|7|11"

# select column you don't want
bracero |>
 select(-date)

# select columns you do want
bracero |>
 select(title, subject, description, date, identifier, language, date_format)

# select columns in a row, except the column you don't want
bracero |>
  select(title:description, identifier:date_format)
```

## rename() function {.smaller}
- sometimes original column names aren't very descriptive, or they're too long. 
- the `rename()` function can help. 
  - let's rename "title" to something more descriptive.
```{r}
#| warning: false
#| message: false
#| echo: true

bracero |>
  rename("interviewee" = "title")
```

## relocate() function {.smaller}
- Say we want the `identifier` column at the very end of the dataframe
- `relocate` automatically puts the column at the front, so you have to specify `.after` or `.before`.
```{r}
#| warning: false
#| message: false
#| echo: true

bracero |> 
 relocate(identifier, .after = language)
```

## Put it together {.smaller}
- We can put all of these data transformation actions together using the pipe `|>` function
```{r}
#| warning: false
#| message: false
#| echo: true

bracero_clean <- bracero |>
  mutate(date_format = as.Date(date)) |>
  select(-date) |>
  rename("interviewee" = "title") |>
  relocate(identifier, .after = date_format)

head(bracero_clean)
```

# Exercise One
- Starting with the `bracero` dataframe, code the following three actions using the `|>` into a new dataframe called `bracero_ex1`: 

  - mutate the `date` column to be in a DD-MM-YYYY format called `date_format`
    - hint: you may need to use the `?as.Date()` helper function
  - select the `title` , `date_format`, and `description` columns
  - reorder the remaining columns by date

# Transforming Data (Rows)
- Moving on to rows! 
  - `filter()` is a function that allows you to keep rows of interest (but won't change the order of your rows) 
  - `arrange()` changes the order of your rows (but won't filter for anything)

## filter() function {.smaller}
- Use the filter() function to select rows that correspond to values of interest. 
- Say we're interested in only interviews that are spoken in spanish:
- use `==` to filter for a direct match 
```{r}
#| warning: false
#| message: false
#| echo: true
bracero_clean |>
  filter(language == "spa")
```

## filter() (cont.) {.smaller}
- You can also combine multiple conditions: 
  - `>` = greater than 
  - `>=` = greater than or equal to
  - `<` = less than
  - `<=` less than or equal to
  - `==` equal to
  - `!=` does not equal

  - Can use `&` to combine conditions
  - `|` for or 
- Perhaps we're interested in interviews that took place after 2004 in English:
```{r}
#| warning: false
#| message: false
#| echo: true
#| eval: false

bracero_clean |>
  filter(date_format >= 2004-01-01 & language == "eng")
```

## filter() and str_detect()
- We can also combine filter with str_detect()
```{r}
#| warning: false
#| message: false
#| echo: true
bracero_clean |> 
  filter(str_detect(interviewee, "Alberto"))
```

## arrange()
- `arrange()` sorts your rows based on the values in the columns. 
- If you sort more than one row, it will sort the first variable, and the others will be used to "break ties"
- including `desc()` wills sort your column values big to small. 
```{r}
#| warning: false
#| message: false
#| echo: true
#| eval: false

bracero_clean |> 
  arrange(date_format, interviewee)
```

# Exercise Two {.smaller}
1. Filter the `bracero_clean` dataframe for any interviews that took place before 2005 with an interviewee whose name includes "Guadalupe".

2. Sort the `bracero_clean` data frame in descending order by date. What was the date, how many interviews took place, and who is listed first?

3. Filter the `bracero_clean` dataframe for any interviewees that mention "California". How many interviewees mention California? 
  - If you're interested in reading descriptions, you can type "bracero_clean$description[]" with the row number in brackets to print the description in your environment window. 


# Grouping Data 

## group_by() function {.smaller}
- `group_by()` function is really helpful for dividing your data into groups for analysis. 
- `group_by()` function does not change your data, but will change subsequent operations if you use them in tandem with a `group_by()`
```{r}
#| warning: false
#| message: false
#| echo: true
#| eval: true

bracero_clean |> 
  group_by(language)
```


## summarize()
- `summarize()` is often used in tandem with the `group_by()` function to calculate a summary statistic for a a group. 
```{r}
#| warning: false
#| message: false
#| echo: true
#| eval: true
bracero_clean |>
 group_by(language) |>
 summarize(n_interviews = n())
```

## slice {.smaller}
- `slice_` functions help select a subset of rows
 - slice_head(n = 5) will take the first five rows of the data
 - slice_tail(n = 5) will take the last five rows of the data
 - slice_min(x, n = 5) will take the five rows of data with the lowest values for column x. 
 - slice_max(x, n = 5) will take the five rows of data with the highest values for column x. 
 - slice_sample(n = 5) will select 5 random rows of data.
```{r}
#| warning: false
#| message: false
#| echo: true
#| eval: true
bracero |>
  slice_sample(n = 15)
```

# Exercise Three {.smaller}
1. How many Braceros were interviewed in this dataset (i.e., Bracero is the subject)? Use a `group_by` function to count.
  - Hint: R is case-sensitive. 
2. How many non-braceros were interviewed in this dataset? 
3.  Using the subject column, can you create a new factor column called `topic_group`that groups the subjects into 3-5 factors?
  - Hint: use `mutate()`, and `case_match()` to classify the groups into a new column called `topic_group`, I've given you a start!
4. Are there any interesting patterns in language spoken among your `topic_groups`? Use a `group_by` and `summarize` function to calculate.

# Exercise Four 
1. Explore the Bracero Archives and listen to at least one interview. 
  - Note that most all bracero interviews are conducted in spanish, if you're interested in a perspective from an english-speaking bracero, I would suggest the interview with [Mr. Cris Luna](https://braceroarchive.org/items/show/138), a Bracero from Ozuluama, Veracruz, Mexico.

  Answer questions 2 and 3 using this week's readings. 