---
title: "[Assess] Bracero Farmworker Case Study"
subtitle: "📍 Coachella Valley, CA"
date: "September 19, 2025"
format: 
  revealjs:
    theme:
      - default
    css: custom.scss
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    center: true
execute:
  freeze: auto
---

# Class Overview 
- Centering activity
- Setting the stage
- Paired coding
- Group Debrief

# Centering Activity

# Setting the Stage
- This week's focus is on data tidying and transformation
- Using metadata from 90 interviews in the Bracero Archive that I scraped
- Important to note that each row represents a person, a life, and a family impacted by the Bracero Program

# Let's get Started
- Find your pair
- Open Posit

# First Steps
- Import Data
- Load packages
```{r}
#| warning: false
#| message: false
#| echo: true

library(readxl) # for loading data
library(tidyverse) # for tidying
```

```{r}
# | eval: true
# | echo: false
bracero <- read_excel("~/Google Drive/My Drive/Teaching/ES-202/data/week_03/bracero_pull.xlsx")
```

# Part II: Today's Goal
- Understand data structure
- Transform and tidy data for necessary questions
- Explore and summarize patterns
- Connect it to theoretical understandings of procedural justice

## Our data {.smaller}
- 90 rows (interviews)
- 6 variables (all `chr` type)
  - title = Person interviewed
  - subject = topic of interview
  - description = summary of interview
  - date = date of interview
  - identifier = link to original interview file
  - language = language interview was conducted in
```{r}
#| warning: false
#| message: false
#| echo: true
glimpse(bracero)
```

## Data Types {.smaller}
- `chr` is character
- `dbl` is double (number)
- `int` is an integer
- `dttm` is date-time 
- `fct` is factor

<span style="color:red;">Thoughts on what variables we may want to change the data types of?</span>

```{r}
#| warning: false
#| message: false
#| echo: false

glimpse(bracero)
```

# Transforming Data (Columns) {.smaller}
- Four functions can change the columns of your data without affecting your rows 
  - mutate(): creates new columns based on existing ones
  - select(): chooses what columns you want in your dataframe
  - rename(): changes the names of current columns
  - relocate(): changes the position of those columns 

## mutate() function {.smaller}
- We use the mutate function to change the `date` column to a date format. 

```{r}
#| warning: false
#| message: false
#| echo: true

bracero |>
  mutate(date_format = as.Date(date)) 
```

## mutate() function (cont.) {.smaller}
- We can also use mutate for creating new columns, and we can pair it with additional functions, such as ifelse statements. 
- Perhaps we were interested in creating a new column that defines whether an interview is from a bracero worker or not. 
  - Note that if we wanted to further classify the subject, we might create a factor column and use the `case_match()` function, rather than an ifelse statement.
```{r}
#| warning: false
#| message: false
#| echo: true 

bracero |> 
  mutate(subject_binary = ifelse(subject %in% c("bracero", "Bracero"), "Bracero", "Non-Bracero" )) 
```

## select() function {.smaller}
- Perhaps we'd like to remove the original `date` column from our dataframe. 
- `select()` can be used to select or deselect columns. 
```{r}
#| warning: false
#| message: false
#| echo: true
#| eval: false
#| code-line-numbers: "|3|7|11"

# select column you don't want
bracero |>
 select(-date)

# select columns you do want
bracero |>
 select(title, subject, description, date, identifier, language, date_format)

# select columns in a row, except the column you don't want
bracero |>
  select(title:description, identifier:date_format)
```

## rename() function {.smaller}
- sometimes original column names aren't very descriptive, or they're too long. 
- the `rename()` function can help. 
  - let's rename "title" to something more descriptive.
```{r}
#| warning: false
#| message: false
#| echo: true

bracero |>
  rename("interviewee" = "title")
```

## relocate() function {.smaller}
- Say we want the `identifier` column at the very end of the dataframe
- `relocate` automatically puts the column at the front, so you have to specify `.after` or `.before`.
```{r}
#| warning: false
#| message: false
#| echo: true

bracero |> 
 relocate(identifier, .after = language)
```

## Put it together {.smaller}
- We can put all of these data transformation actions together using the pipe `|>` function
```{r}
#| warning: false
#| message: false
#| echo: true

bracero_clean <- bracero |>
  mutate(date_format = as.Date(date)) |>
  select(-date) |>
  rename("interviewee" = "title") |>
  relocate(identifier, .after = date_format)

head(bracero_clean)
```

# Exercise One
- Starting with the `bracero` dataframe, code the following three actions using the `|>` into a new dataframe: 

  - mutate the `date` column to be in a DD-MM-YYYY format called `date_format`
    - hint: you may need to use the `?as.Date()` helper function
  - select the `title` , `date_format`, and `description`columns
  - reorder the remaining columns by date

# Transforming Data (Rows)
- Moving on to rows! 
  - `filter()` is a function that allows you to keep rows of interest (but won't change the order of your rows) 
  - `arrange()` changes the order of your rows (but won't filter for anything)

## filter() function 
- Use the filter() function to select rows that correspond to values of interest. 
- Say we're interested in only interviews that are spoken in spanish:
- use `==` to filter for a direct match 
```{r}
#| warning: false
#| message: false
#| echo: true
bracero_clean |>
  filter(language == "spa")
```

## filter() (cont.) {.smaller}
- You can also combine multiple conditions: 
  - `>` = greater than 
  - `>=` = greater than or equal to
  - `<` = less than
  - `<=` less than or equal to
  - `==` equal to
  - `!=` does not equal

  - Can use `&` to combine conditions
  - `|` for or 
- Perhaps we're interested in interviews that took place after 2004 in English:
```{r}
#| warning: false
#| message: false
#| echo: true

bracero_clean |>
  filter(date_format >= 2004-01-01 & language == "eng")
```

## filter() and str_detect()
- We can also combine filter with str_detect()
```{r}
#| warning: false
#| message: false
#| echo: true
bracero_clean |> 
  filter(str_detect(interviewee, "Alberto"))
```

## arrange()
- `arrange()` sorts your rows based on the values in the columns. 
- If you sort more than one row, it will sort the first variable, and the others will be used to "break ties"
- including `desc()` wills sort your column values big to small. 
```{r}
#| warning: false
#| message: false
#| echo: true

bracero_clean |> 
  arrange(date_format, interviewee)
```

# Exercise Two
1. Filter the `bracero_clean` dataframe for any interviews that took place before 2005 with an interviewee whose name includes "Guadalupe".

2. Sort the `bracero_clean` data frame in descending order by date. What was the date, how many interviews took place, and who is listed first?

3. Filter the `bracero_clean` dataframe for any interviewees that mention "California". How many interviewees mention California? 
  - If you're interested in reading descriptions, you can type "bracero_clean$description[]" with the row number in brackets to print the description in your environment window. 


# Grouping Data 

## group_by() function


## summarize()

# Exercise Three
1. How many Braceros were interviewed in this dataset (i.e., Bracero is the topic)? Use a `group_by` function to count.
  - Hint: R is case-sensitive. 
2. How many non-braceros were interviewed in this dataset? 
3. Based on the non-bracero topics, can you create a new factor column that groups the topics into 3-5 factors?
  - use `mutate()`, and `case_match()` to classify the groups into a new column called `topic_group`
4. Are there any interesting patterns in language spoken among your `topic_groups`? Use a `group_by` and `summarize` function to calculate.

# Exercise Four 
1. Listen to an [interview with Mr. Cris Luna](https://braceroarchive.org/items/show/138), a Bracero from Ozuluama, Veracruz, Mexico. 

[interview with Mr. Raymundo Villa](https://braceroarchive.org/items/show/81), an interpreter from Ysleta, Texas who worked with 

Explore the Bracero archives, and listen to or read the biographical descriptions of interviews in the "Oral History" section. 
2. Reflect on the aspects of procedural injustice as they relate to othe Bracero program: 
  - 