---
title: "HW 02"
author: "<YOUR NAME HERE>"
date: "`r Sys.Date()`"
format: html
---

# Introduction to Quarto
```{r}
#| label: setup
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)
```

## Quarto and R Markdown

This is a Quarto document. It is a structured plain-text format that allows you to mix text and pieces (also called "chunks" or "cells") of code in a way that can be turned into HTML, PDF, and MS Word documents. For more details on using Quarto see quarto's [getting started page](https://quarto.org/docs/get-started/hello/rstudio.html). 

When you click the **Render** button in RStudio, a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r}
#| label: addition

2 + 2
```

The chunk is delimited by those backticks and curly braces on line 28 (to open it) and the three backticks on line 30 (to close it). You can also use a Mac shortcut (command + option + i) to create an empty chunk. Here is what an empty chunk looks like:
```{r}
#| label: code chunk

```

# HW 02: Solid Waste Sites Case Study
## Learning Outcomes
- Understand what R can do
- Recreate tables 1-3 in Bullard's Study 
- Visualize Distributional Justice
- Connect it to theoretical understandings of distributional justice

# Understanding What R Can do

## Installing Packages
```{r}
#| label: installing packages

```

```{r}
#| label: load packages

```

## Loading data
```{r}
#| label: loading data

```

## Sites Data 
```{r}
#| label: view data
```

# Recreating Tables 1 through 3
## Table one
```{r}
#| label: create table one df
```

```{r}
#| label: format table one
```

## EX_1: Table two
- Recreate tables 2 and 3 using the same approaches
- Pay close attention to the information shared in each table
```{r}
#| label: Repeat for table two
```

## EX_1: Table three
```{r}
#| label: Repeat for table three
```

# Visualize Distributional Justice
## Recreating Table 4
```{r}
#| label: crosstabs table
```

```{r}
#| label: format table
```

## EX_2: One more summary table
- Suppose you lived in a Southwest neighborhood and were interested in the distribution of facility types by Southwest versus non-southwest neighborhoods.
- Create a cross tabs summary table (make sure it's formatted nicely!) 
```{r}
#| label: neighborhood summary table
```

## Plot the data
```{r}
#| label: first ggplot layer

```

## Add Variables
```{r}
#| label: first ggplot layer
```

## Add a Visual
```{r}
#| label: define geom
```

## Format colors and labels
```{r}
#| label: define geom
```

## EX_03: Troubleshooting with a vignette
- Read the vignette for the `scale_fill_manual()` package. What package does it recommend for color blindness?
- type your answer below. 

## EX_04: Connecting to Distributional Justice
- 

