---
title: "Data Sources"
date: "December 09, 2025"
categories: "Final Project"
---
## Variable Overview 
```{r}
#| echo: false
#| eval: true
#| output: false
library(readxl)
library(tidyverse)
library(knitr)
library(kableExtra)

vars <- read_excel("/Users/mc152/Library/CloudStorage/GoogleDrive-mc152@wellesley.edu/My Drive/Teaching/ES-202/Final/Final_Dataset_variable_names.xlsx")
```

```{r}
#| echo: false
#| eval: true
#| output: true

options(knitr.kable.NA = '')
vars %>%
  kable("html", escape = FALSE) %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(
    which(names(vars) == "Description"), 
    width = "30em", 
    extra_css = "word-wrap: break-word; white-space: normal;"
  )
```


## Data Download
The dataset can be downloaded [here](https://drive.google.com/drive/folders/1hZWFZCzqoeWvpHIh0ZIbBVR8hKqKJBfA). 

It can also be viewed as an interactive map [here](https://arcg.is/0CKCDb0). To view different variables: 

 - Click the "layers" tab on the upper left corner
 - Select the "ES-202 Final Shapefile" layer
 - Select the "styles" tab on the upper right corner
 - Choose 1-2 "atttributes" using the shapefile variable names above 
 - Pick a style to map the variables (I like "relationship" for 2 variables, and "counts and amounts" for 1 variable)

